Faithful Variable Screening for High-Dimensional Convex Regression

### 2014-09-11 tex/scam/optimization.tex
modified the order statistic notation

The sparse convex additive model optimization in (\ref{np}) is a quadratic program with
$O(np)$ variables and $O(np)$ constraints. 
Directly applying a QP solver for $f$ and $\beta$
is computationally expensive for relatively large
$n$ and $p$. However, notice that variables in different feature
dimensions are only coupled in the squared error term
$(y_{i}-\mu - \sum_{k=1}^{p}f_{ik})^{2}$. Hence, we can apply the block
coordinate descent method, where in each step we solve the following
QP subproblem for $\{f_k, \beta_k\}$ with the
other variables fixed. In matrix notation, the optimization is
\begin{align}
\begin{split}
\min_{ f_k, \beta_k, \gamma_k} \;\;& \frac{1}{2n} \| \uds{r}_k - \uds{f}_k \|_2^2 
     + \lambda \gamma_k \label{opt:1d_compact} \\
 \textrm{such that } & P_k \uds{f}_k = \diag(P_k \bds{x}_k)  \uds{\beta}_k \\
   & D \uds{\beta}_k \leq 0 \\
   & -\gamma_k \mathbf{1}_n \leq \uds{f}_k \leq \gamma_k \mathbf{1}_n   \\
   & \mathbf{1}_n^\tran \uds{f}_k = 0 
\end{split}
\end{align}
where $\uds{\beta}_k \in \R^{n-1}$ is the vector $\uds{\beta}_k =
(\beta_{1k}, \ldots, \beta_{(n-1)k})^T$, and
$\uds{r}_{k} \in \R^n$ is the residual vector $\uds{r}_{k} = (y_i -
\hat\mu - \sum_{k' \neq k} f_{ik'})^T$.
In addition, 
$P_k \in \R^{(n-1) \times n}$ is a permutation matrix where the $i$-th
row  is all zeros except for the value $-1$ in position $\perm{i}$ and
the value $1$ in
position $\perm{i+1}$, and $D \in \R^{(n-2) \times (n-1)}$ is another
permutation matrix  where the $i$-th row is all zeros except for a $1$ 
in position $i$ and a $-1$ in position $i+1$.  We denote by
$\diag( v )$ the diagonal matrix with diagonal entries $v$.
The extra variable $\gamma_{k}$ is introduced to impose the
regularization penalty involving the $\ell_{\infty}$ norm.

@@

This QP
subproblem involves $O(n)$ variables, $O(n)$ constraints and a sparse
structure, which can be solved efficiently using optimization
packages. In our experiments we use {\sc mosek} (\href{http://www.mosek.com/}{www.mosek.com}).  We cycle through
all covariates $k$ from $1$ to $p$ multiple times until convergence.
Empirically, we observe that the algorithm converges in only a few
cycles. We also implemented an ADMM solver for (\ref{np})
\citep{Boyd:admm}, but found
that it is not as efficient as this QP solver.

After optimization, the function estimates for an input vector $\bds{x}$ is, according to (\ref{hyper}),
\begin{equation}
\begin{split}
      \hat f(\bds{x}) & = \sum_{k=1}^{p} \hat f_k(x_{j})+ \hat \mu 
= \sum_{k=1}^{p}\max_{i} \bigl\{\hat f_{ik}+ \hat \beta_{ik}(x_{k}-x_{ik})\bigr\} +
      \hat \mu.
\end{split}
\end{equation}

The univariate concave function estimation is a straightforward
modification of optimization~\ref{opt:1d_compact}. It is only
necessary to modify the linear inequality constraints so that the subgradients are
non-increasing: $\beta_{\perm{i+1}k} \leq \beta_{\perm{i}k}$.

>>>>

The sparse convex additive model optimization in (\ref{np}) is a quadratic program with
$O(np)$ variables and $O(np)$ constraints. 
Directly applying a QP solver for $f$ and $\beta$
is computationally expensive for relatively large
$n$ and $p$. However, notice that variables in different feature
dimensions are only coupled in the squared error term
$(y_{i}-\mu - \sum_{k=1}^{p}f_{ik})^{2}$. Hence, we can apply the block
coordinate descent method, where in each step we solve the following
QP subproblem for $\{f_k, \beta_k\}$ with the
other variables fixed. In matrix notation, the optimization is
\begin{align}
\begin{split}
\min_{ f_k, \beta_k, \gamma_k} \;\;& \frac{1}{2n} \| \uds{r}_k - \uds{f}_k \|_2^2 
     + \lambda \gamma_k \label{opt:1d_compact} \\
 \textrm{such that } & P_k \uds{f}_k = \diag(P_k \bds{x}_k)  \uds{\beta}_k \\
   & D \uds{\beta}_k \leq 0 \\
   & -\gamma_k \mathbf{1}_n \leq \uds{f}_k \leq \gamma_k \mathbf{1}_n   \\
   & \mathbf{1}_n^\tran \uds{f}_k = 0 
\end{split}
\end{align}
where $\uds{\beta}_k \in \R^{n-1}$ is the vector $\uds{\beta}_k =
(\beta_{1k}, \ldots, \beta_{(n-1)k})^T$, and
$\uds{r}_{k} \in \R^n$ is the residual vector $\uds{r}_{k} = (y_i -
\hat\mu - \sum_{k' \neq k} f_{ik'})^T$.
In addition, 
$P_k \in \R^{(n-1) \times n}$ is a permutation matrix where the $i$-th
row  is all zeros except for the value $-1$ in position $\perm{i}$ and
the value $1$ in
position $\perm{i+1}$, and $D \in \R^{(n-2) \times (n-1)}$ is another
permutation matrix  where the $i$-th row is all zeros except for a
value $1$  in position $\perm{i}$ and a value $-1$ in position $\perm{i+1}$.  We denote by
$\diag( v )$ the diagonal matrix with diagonal entries $v$.
The extra variable $\gamma_{k}$ is introduced to impose the
regularization penalty involving the $\ell_{\infty}$ norm.

@@

This QP
subproblem involves $O(n)$ variables, $O(n)$ constraints and a sparse
structure, which can be solved efficiently using optimization
packages. In our experiments we use {\sc mosek} (\href{http://www.mosek.com/}{www.mosek.com}).  We cycle through
all covariates $k$ from $1$ to $p$ multiple times until convergence.
Empirically, we observe that the algorithm converges in only a few
cycles. We also implemented an ADMM solver for (\ref{np})
\citep{Boyd:admm}, but found
that it is not as efficient as this blockwise QP solver.

After optimization, the function estimate for an input vector $\bds{x}$ is, according to (\ref{hyper}),
\begin{equation}
\begin{split}
      \hat f(\bds{x}) & = \sum_{k=1}^{p} \hat f_k(x_{j})+ \hat \mu 
= \sum_{k=1}^{p}\max_{i} \bigl\{\hat f_{ik}+ \hat \beta_{ik}(x_{k}-x_{ik})\bigr\} +
      \hat \mu.
\end{split}
\end{equation}

The univariate concave function estimation required in the DC stage is a straightforward
modification of optimization~\eqref{opt:1d_compact}. It is only
necessary to modify the linear inequality constraints so that the subgradients are
non-increasing: $\beta_{\perm{i+1}k} \leq \beta_{\perm{i}k}$.
